Northern Pikeminnow Abundance in Deadwater Slough, Salmon River, Idaho, and Potential Impacts to Local Chinook Salmon Populations

### 2022-03-30 analysis/paper/paper.Rmd
made sure all numbers in text jived

Predation on emigrating juvenile salmonids by piscivorous fishes is a widely-studied source of mortality within reservoirs and dam tailraces in the Columbia River basin. Native Northern Pikeminnow have been estimated to consume ~8% of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia rivers, annually. Less is known about the interaction between upstream, river-dwelling Northern Pikeminnow and their impacts on salmonid recovery above the reservoir systems. In this study, we examine the abundance of Northern Pikeminnow in a slow-water reach of the Salmon River, Idaho, known as Deadwater Slough, and consider their potential impacts on local Chinook Salmon populations. The Deadwater Slough contains historically important overwinter rearing habitat and is a migration corridor for Chinook Salmon as well as other Endangered Species Act-listings including steelhead and Sockeye Salmon. We estimated the abundance of Northern Pikeminnow in Deadwater Slough to be 27,874 in the fall and 19,499 in the spring, corresponding with the peak emigrations of juvenile Chinook Salmon. Using these abundance values, we estimated Northern Pikeminnow consumption of juvenile Chinook Salmon. Assuming 60% of the Northern Pikeminnow diet is fish, of which 50% is juvenile Chinook Salmon, we estimated the Northern Pikeminnow population can consume 95,120 juvenile Chinook Salmon, annually. Using a sensitivity analysis, Northern Pikeminnow predation on juvenile Chinook Salmon was predicted to reduce adult returns by 539 (95% CI: 231 - 1,335). Given the relatively small size of the Salmon River, the large densities and potential consumption rates of Northern Pikeminnow in Deadwater Slough suggest predation likely has consequential impacts on Chinook Salmon in the Upper Salmon River major population group.

@@

spring_abund = all_est %>%
  filter(model == "Schnabel") %>%
  #filter(model == "Schnabel - Delayed Mixing") %>%
  select(-range, -prop_range) %>%
  inner_join(total_cpue,
             by = c("event_name", "species")) %>%
  select(species,
         model,
         model_event = event_name,
         N_F = N,
         Lci_F = Lci,
         Uci_F = Uci,
         cpue_F = catch_per_hr) %>%
  group_by(species, model) %>%
  summarise(across(c(N_F:cpue_F),
                   mean),
            .groups = "drop") %>%
    full_join(total_cpue %>%
              filter(str_detect(event_name, "Spring")) %>%
              select(species,
                     cpue_S = catch_per_hr),
            by = "species") %>%
  # calculate spring abundance estimate
  mutate(N_S = N_F * (cpue_S / cpue_F),
         Lci_S = Lci_F * (cpue_S / cpue_F),
         Uci_S = Uci_F * (cpue_S / cpue_F))

@@

Mark-recapture abundance estimates of Northern Pikeminnow in Deadwater Slough ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` in fall 2019 and from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` in fall 2020 (Table \@ref(tab:abund-tab)). We estimated larger populations of Northern Pikeminnow using the unadjusted multiple-census estimator compared to adjusted and single census estimators (Figure \@ref(fig:abund-fig)). Because our sampling design most closely matched a multiple census estimator and required a small sample size modification, we deemed the Schnabel estimates most appropriate. Accordingly, the mean Northern Pikeminnow abundance for the two fall sampling events was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel"]), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_F, spring_abund$Uci_F)), big.mark = ","), collapse = " - ")`). Using the delayed-mixing Schnabel estimator, the mean fall abundance estimate was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` (95% CI: `r prettyNum(mean(all_est$Lci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` - `r prettyNum(mean(all_est$Uci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")`). For spring 2021, we estimated `r prettyNum(round(spring_abund$N_S), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_S, spring_abund$Uci_S)), big.mark = ","), collapse = " - ")`) Northern Pikeminnow in Deadwater Slough. Those estimates translate to linear densities of `r prettyNum(round(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms), big.mark = ",")` and `r prettyNum(round(spring_abund$N_S / ddw_rkms), big.mark = ",")` Northern Pikeminnow per rkm and areal densities (fish/100 m$^2$) of `r round((mean(all_est$N[all_est$model == "Schnabel"]) / ddw_m2) * 100, 1)` and `r round((spring_abund$N_S / ddw_m2) * 100, 1)` for fall and spring, respectively.

@@

We found that gastric lavage successfully removed food items in all dissected individuals, indicating the method was effective. Of the `r prettyNum(gastric_tbl$lavaged[gastric_tbl$event_name == "Total"], big.mark = ",")` Northern Pikeminnow sampled using gastric lavage, we found contents in `r prettyNum(gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` (`r prettyNum((gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Total"]) * 100, digits = 3)`%) stomachs and confirmed fish or fish parts in `r prettyNum(gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` (`r prettyNum((gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"]) * 100, digits = 3)`%) stomachs. Northern Pikeminnow captured during spring 2021 had a higher proportion of fish content in their diet compared to individuals captured in the fall surveys (Table \@ref(tab:gastric-tab)). The mean wet weight of total contents for an individual was `r prettyNum(mean(stmch_df$stomach_contents_weight), digits = 2)` g (median = `r prettyNum(median(stmch_df$stomach_contents_weight), digits = 3)` g; SD = `r round(sd(stmch_df$stomach_contents_weight), 2)` g). Overall, fish or fish remnants made up `r prettyNum(mean(stmch_df$p_fish) * 100, digits = 3)`% of all stomach contents examined.

@@

```{r tons-consumption}
tot_g_DSR = (bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f60_g"] * mean(all_est$N[all_est$model == "Schnabel"])) / 1000000
tot_g_NRR = (bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f60_g"] * round(spring_abund$N_S)) / 1000000

@@

During the fall DSR emigration, we estimate an average-size Northern Pikeminnow (`r round(avg_fall_NPM_TL, 1)` TL mm) to consume `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f60_g"]` g of fish to maintain their body size, assuming 60% of their diet consists of fish prey (Figure \@ref(fig:bio-p)). Fall consumption ranged from `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f30_g"]` g for a diet of 30% fish to `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f90_g"]` g for a diet of 90% fish. During the spring NRR emigration, we estimated an average-size Northern Pikeminnow (`r round(avg_spring_NPM_TL, 1)` TL mm) to consume `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f60_g"]` g of fish, assuming 60% of their diet is fish prey (Figure \@ref(fig:bio-p)). Spring consumption ranged from `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f30_g"]` for a diet of 30% fish to `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f90_g"]` for a diet of 90% fish. Given estimated mean Northern Pikeminnow population sizes of `r prettyNum(mean(all_est$N[all_est$model == "Schnabel"]), big.mark = ",")` in the fall, `r prettyNum(round(spring_abund$N_S), big.mark = ",")` in the spring 2021, and a diet consisting of 60% fish, the Northern Pikeminnow population can consume `r round(tot_g_DSR, 1)` metric tons of fish during the fall DSR emigration and `r round(tot_g_NRR, 1)` metric tons during the spring NRR emigration in Deadwater Slough.

@@

All mark-recapture abundance estimators suggest a large population of Northern Pikeminnow occupies Deadwater Slough. The mean population size of Northern Pikeminnow was estimated to be `r prettyNum(mean(all_est$N[all_est$model == "Schnabel"]), big.mark = ",")` during the fall DSR emigration and `r prettyNum(round(spring_abund$N_S), big.mark = ",")` during the spring NRR emigration. Those estimates translate to linear densities (fish/rkm) that are `r round((mean(all_est$N[all_est$model == "Schnabel"]) / ddw_rkms) / john_day_dens)` and `r round((spring_abund$N_S / ddw_rkms) / john_day_dens)` times greater than densities reported by @Beamesderfer1991 for the John Day Reservoir and approximately `r round((mean(all_est$N[all_est$model == "Schnabel"]) / ddw_rkms) / low_columbia_fish_km)` and `r round((spring_abund$N_S / ddw_rkms) / low_columbia_fish_km)` times greater than projections for the lower Columbia River [@Beamesderfer1996]. @Beamesderfer1991 acknowledge that their methods were unsuitable for sampling offshore, noting that water velocity, depth, irregular bottom contours, and barge traffic made sampling offshore ineffective; therefore it is possible that their estimates may have been low. Conversely, at Deadwater Slough the maximum depth was approximately 6 m with a relatively homogenous bottom contour, making angling an effective method throughout the entire reach. Our estimated densities of Northern Pikeminnow suggest that slow-water reaches outside of the reservoir complexes on the Snake and Columbia rivers may support exceptionally high predator densities, consistent with findings of @Harnish2014, @Gray2001, and @Zimmerman1999. Given the dearth of current data available in the Columbia River basin, it is unclear how Northern Pikeminnow abundance estimates in this study compare to elsewhere in the Columbia River Basin; contemporary information on piscine predator abundances may be needed [@Widener2021].

>>>>

Predation on emigrating juvenile salmonids by piscivorous fishes is a widely-studied source of mortality within reservoirs and dam tailraces in the Columbia River basin. Native Northern Pikeminnow have been estimated to consume ~8% of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia rivers, annually. Less is known about the interaction between upstream, river-dwelling Northern Pikeminnow and their impacts on salmonid recovery above the reservoir systems. In this study, we examine the abundance of Northern Pikeminnow in a slow-water reach of the Salmon River, Idaho, known as Deadwater Slough, and consider their potential impacts on local Chinook Salmon populations. The Deadwater Slough contains historically important overwinter rearing habitat and is a migration corridor for Chinook Salmon as well as other Endangered Species Act-listings including steelhead and Sockeye Salmon. We estimated the abundance of Northern Pikeminnow in Deadwater Slough to be 19,499 in the fall and 10,352 in the spring, corresponding with the peak emigrations of juvenile Chinook Salmon. Using these abundance values, we estimated Northern Pikeminnow consumption of juvenile Chinook Salmon. Assuming 60% of the Northern Pikeminnow diet is fish, of which 50% is juvenile Chinook Salmon, we estimated the Northern Pikeminnow population can consume 61,409 juvenile Chinook Salmon, annually. Using a sensitivity analysis, Northern Pikeminnow predation on juvenile Chinook Salmon was predicted to reduce adult returns by 377 (95% CI: 161 - 935). Given the relatively small size of the Salmon River, the large densities and potential consumption rates of Northern Pikeminnow in Deadwater Slough suggest predation likely has consequential impacts on Chinook Salmon in the Upper Salmon River major population group.

@@

spring_abund = all_est %>%
  filter(model == "Schnabel - Delayed Mixing") %>%
  # filter(model == "Schnabel") %>%
  select(-range, -prop_range) %>%
  inner_join(total_cpue,
             by = c("event_name", "species")) %>%
  select(species,
         model,
         model_event = event_name,
         N_F = N,
         Lci_F = Lci,
         Uci_F = Uci,
         cpue_F = catch_per_hr) %>%
  group_by(species, model) %>%
  summarise(across(c(N_F:cpue_F),
                   mean),
            .groups = "drop") %>%
    full_join(total_cpue %>%
              filter(str_detect(event_name, "Spring")) %>%
              select(species,
                     cpue_S = catch_per_hr),
            by = "species") %>%
  # calculate spring abundance estimate
  mutate(N_S = N_F * (cpue_S / cpue_F),
         Lci_S = Lci_F * (cpue_S / cpue_F),
         Uci_S = Uci_F * (cpue_S / cpue_F))

@@

Mark-recapture abundance estimates of Northern Pikeminnow in Deadwater Slough ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` in fall 2019 and from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` in fall 2020 (Table \@ref(tab:abund-tab)). We estimated larger populations of Northern Pikeminnow using the unadjusted multiple-census estimator compared to adjusted and single census estimators (Figure \@ref(fig:abund-fig)). Because our sampling design most closely matched a multiple census estimator and required a small sample size modification, we deemed the Schnabel estimates most appropriate. Accordingly, the mean Northern Pikeminnow abundance for the two fall sampling events was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel"]), big.mark = ",")` (95% CI: `r prettyNum(round(mean(all_est$Lci[all_est$model == "Schnabel"])), big.mark = ",")` - `r prettyNum(round(mean(all_est$Uci[all_est$model == "Schnabel"])), big.mark = ",")`). Using the delayed-mixing Schnabel estimator, the mean fall abundance estimate was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` (95% CI: `r prettyNum(mean(all_est$Lci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` - `r prettyNum(mean(all_est$Uci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")`). For spring 2021, we estimated `r prettyNum(round(spring_abund$N_S), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_S, spring_abund$Uci_S)), big.mark = ","), collapse = " - ")`) Northern Pikeminnow in Deadwater Slough. Those estimates translate to linear densities of `r prettyNum(round(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms), big.mark = ",")` and `r prettyNum(round(spring_abund$N_S / ddw_rkms), big.mark = ",")` Northern Pikeminnow per rkm and areal densities (fish/100 m$^2$) of `r round((mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_m2) * 100, 1)` and `r round((spring_abund$N_S / ddw_m2) * 100, 1)` for fall and spring, respectively.

@@

We found that gastric lavage successfully removed food items in all dissected individuals, indicating the method was effective. Of the `r prettyNum(gastric_tbl$lavaged[gastric_tbl$event_name == "Total"], big.mark = ",")` Northern Pikeminnow sampled using gastric lavage, we found contents in `r prettyNum(gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` (`r prettyNum((gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Total"]) * 100, digits = 3)`%) stomachs and confirmed fish or fish parts in `r prettyNum(gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` stomachs. Northern Pikeminnow captured during spring 2021 had a higher proportion of fish content in their diet compared to individuals captured in the fall surveys (Table \@ref(tab:gastric-tab)). The mean wet weight of total contents for an individual was `r prettyNum(mean(stmch_df$stomach_contents_weight), digits = 2)` g (median = `r prettyNum(median(stmch_df$stomach_contents_weight), digits = 3)` g; SD = `r round(sd(stmch_df$stomach_contents_weight), 2)` g). Overall, fish or fish remnants made up `r prettyNum(mean(stmch_df$p_fish) * 100, digits = 3)`% of all stomach contents examined.

@@

```{r tons-consumption}
tot_g_DSR = (bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f60_g"] * mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"])) / 1000000
tot_g_NRR = (bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f60_g"] * round(spring_abund$N_S)) / 1000000

@@

During the fall DSR emigration, we estimate an average-size Northern Pikeminnow (`r round(avg_fall_NPM_TL, 1)` TL mm) to consume `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f60_g"]` g of fish to maintain their body size, assuming 60% of their diet consists of fish prey (Figure \@ref(fig:bio-p)). Fall consumption ranged from `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f30_g"]` g for a diet of 30% fish to `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f90_g"]` g for a diet of 90% fish. During the spring NRR emigration, we estimated an average-size Northern Pikeminnow (`r round(avg_spring_NPM_TL, 1)` TL mm) to consume `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f60_g"]` g of fish, assuming 60% of their diet is fish prey (Figure \@ref(fig:bio-p)). Spring consumption ranged from `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f30_g"]` for a diet of 30% fish to `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f90_g"]` for a diet of 90% fish. Given estimated mean Northern Pikeminnow population sizes of `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` in the fall, `r prettyNum(round(spring_abund$N_S), big.mark = ",")` in the spring 2021, and a diet consisting of 60% fish, the Northern Pikeminnow population can consume `r round(tot_g_DSR, 1)` metric tons of fish during the fall DSR emigration and `r round(tot_g_NRR, 1)` metric tons during the spring NRR emigration in Deadwater Slough.

@@

All mark-recapture abundance estimators suggest a large population of Northern Pikeminnow occupies Deadwater Slough. The mean population size of Northern Pikeminnow was estimated to be `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` during the fall DSR emigration and `r prettyNum(round(spring_abund$N_S), big.mark = ",")` during the spring NRR emigration. Those estimates translate to linear densities (fish/rkm) that are `r round((mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms) / john_day_dens)` and `r round((spring_abund$N_S / ddw_rkms) / john_day_dens)` times greater than densities reported by @Beamesderfer1991 for the John Day Reservoir and approximately `r round((mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms) / low_columbia_fish_km)` and `r round((spring_abund$N_S / ddw_rkms) / low_columbia_fish_km)` times greater than projections for the lower Columbia River [@Beamesderfer1996]. @Beamesderfer1991 acknowledge that their methods were unsuitable for sampling offshore, noting that water velocity, depth, irregular bottom contours, and barge traffic made sampling offshore ineffective; therefore it is possible that their estimates may have been low. Conversely, at Deadwater Slough the maximum depth was approximately 6 m with a relatively homogenous bottom contour, making angling an effective method throughout the entire reach. Our estimated densities of Northern Pikeminnow suggest that slow-water reaches outside of the reservoir complexes on the Snake and Columbia rivers may support exceptionally high predator densities, consistent with findings of @Harnish2014, @Gray2001, and @Zimmerman1999. Given the dearth of current data available in the Columbia River basin, it is unclear how Northern Pikeminnow abundance estimates in this study compare to elsewhere in the Columbia River Basin; contemporary information on piscine predator abundances may be needed [@Widener2021].
